Classless Java

### 2016-03-24 OOPSLA16/sections/Conclusion.tex
please use these commands in further editions to unify hyphen usage: \objectoriented \interfacebased \classbased \prototypebased ... you may refer to file predef/def.tex

Before Java 8, concrete methods and static methods were not allowed
to appear in interfaces.  Java 8 allows static interface methods and
introduces \emph{default methods}, which enables implementations
inside interfaces. This had an important positive consequence that
was probably overlooked: the concept of class
(in Java) is now (almost) redundant and unneeded.
We proposed a programming style, called ClassLess Java, where
truly object oriented programs and (reusable) libraries
can be defined and used without ever defining a single class.

@@

Class less Java is just a programming style, but is 
showing the way for a new flavour of object orientation:
We propose Interface Based Object Oriented languages (IB),
as opposed to Class Based or Prototype Based.

>>>>

Before Java 8, concrete methods and static methods were not allowed
to appear in interfaces.  Java 8 allows static interface methods and
introduces \emph{default methods}, which enables implementations
inside interfaces. This had an important positive consequence that
was probably overlooked: the concept of class
(in Java) is now (almost) redundant and unneeded.
We proposed a programming style, called ClassLess Java, where
truly \objectoriented programs and (reusable) libraries
can be defined and used without ever defining a single class.

@@

Class less Java is just a programming style, but is 
showing the way for a new flavour of object orientation:
We propose \interfacebased \objectoriented languages (IB),
as opposed to \classbased or \prototypebased.

### 2016-03-24 OOPSLA16/sections/FutureWork.tex
please use these commands in further editions to unify hyphen usage: \objectoriented \interfacebased \classbased \prototypebased ... you may refer to file predef/def.tex

This can raise questions: keeping fields private is a staple of CB object oriented programming;
by forcing the user/heir to rely only on methods, we avoid (a part)
of the fragile base class problem and we obtain representation independence.

>>>>

This can raise questions: keeping fields private is a staple of CB \objectoriented programming;
by forcing the user/heir to rely only on methods, we avoid (a part)
of the fragile base class problem and we obtain representation independence.

### 2016-03-24 OOPSLA16/sections/Introduction.tex
please use these commands in further editions to unify hyphen usage: \objectoriented \interfacebased \classbased \prototypebased ... you may refer to file predef/def.tex

Object oriented languages strive to offer great code reuse.
They aim to couple flexibility and rigour, expressive power and
modular reasoning.  Two main ideas emerged to this end: prototype
based (PB)~\cite{Ungar87self} and class based (CB) languages such as
Java or Scala.  In prototype based
languages objects inherit from other objects. Thus objects own
both behaviour and state (and objects are all you have).
In class based languages an object is an instance of a specific class,
and classes inherit from other classes.  Here objects own the state,
while classes contain behaviour and the structure of the state.

This paper presents a third alternative: the concept of
\textbf{Interface based} (IB) object oriented languages, where objects
implement interfaces directly. In IB interfaces own the implementation
for the behaviour, which is structurally defined in their
interface. Programmers do not define objects directly, but delegate
the task to \emph{object interfaces}, whose role is similar to non-abstract 
classes in CB languages. Instantiation of objects is
done using static factory methods in object interfaces.

>>>>

\Objectoriented languages strive to offer great code reuse.
They aim to couple flexibility and rigour, expressive power and
modular reasoning.  Two main ideas emerged to this end: prototype
based (PB)~\cite{Ungar87self} and \classbased (CB) languages such as
Java or Scala.  In \prototypebased
languages objects inherit from other objects. Thus objects own
both behaviour and state (and objects are all you have).
In \classbased languages an object is an instance of a specific class,
and classes inherit from other classes.  Here objects own the state,
while classes contain behaviour and the structure of the state.

This paper presents a third alternative: the concept of
\textbf{\interfacebased} (IB) \objectoriented languages, where objects 
implement interfaces directly. In IB interfaces own the implementation
for the behaviour, which is structurally defined in their
interface. Programmers do not define objects directly, but delegate
the task to \emph{object interfaces}, whose role is similar to non-abstract 
classes in CB languages. Instantiation of objects is
done using static factory methods in object interfaces.

### 2016-03-24 OOPSLA16/sections/RelatedWork.tex
please use these commands in further editions to unify hyphen usage: \objectoriented \interfacebased \classbased \prototypebased ... you may refer to file predef/def.tex

\paragraph{Models of Object-Oriented programming}
There are traditionally two big styles of OOP: class-based and
prototype-based. The class-based tradition started with Simula~\cite{Dahl66simula},
and is the dominant style of OOP today. Major languages such
as Java, C\#, C++ or Scala favour a class based OOP style. Very often
class-based languages are statically typed and they have interfaces or
some other similar mechanisms such as traits. However, in contrast to IB
programming, interfaces or traits in CB programming are not meant to
replace classes, but rather, they complement classes. Normally
interfaces or traits define the signatures of the methods and
\emph{may} have some behaviour (such as in Java 8 interfaces or Scala
traits). However classes are still responsible for object
instantiation and handling state. Prototype-based languages started
with Self~\cite{Ungar87self}, and are still widely used today via languages like
JavaScript. Frequently prototype-based languages are dynamically
typed. IB programming is clearly closer to CB programming. However the
key difference is that classes are not needed in IB
programming. Object interfaces recover the responsibility of object
instantiation, and abstract state operations provide a way to deal
with state. A primary motivation of IB programming is to improve
modularity and flexibility. Central to these goals are the
more expressive and automated forms of type-refinement, and support
for multiple interface inheritance.

\paragraph{Multiple inheritance in Object-Oriented languages}
Many authors have argued in favor or against multiple inheritance.
Multiple \emph{implementation}
inheritance is very expressive, but difficult to model and implement, and can
cause difficulties (including the famous diamond (fork-join)
problem~\cite{bracha90mixin,Sak89dis}, conflicting methods, etc.) in reasoning about
programs. To conciliate the need for expressive power and simplicity, many
models have been proposed, including C++ virtual inheritance,
mixins~\cite{bracha90mixin}, traits~\cite{scharli03traits}, and hybrid models
such as CZ~\cite{malayeri2009cz}.  They provide novel programming architecture
models in the OO paradigm. In terms of restrictions set on these models, C++
virtual inheritance aims at a relative general model; the mixin model adds some
restrictions; and the trait model is the most restricted one (excluding
state, instantiation, etc).

@@

\paragraph{ThisType and MyType}
%\marco{ this section is unfair. There are unsolved problems with ThisType in negative positions, namely it breaks subtyping. The most common solution is to allow calling some methods only when the exact type is known. This demolishes most advantages of interfaces.}
Object interfaces support automatic type-refinement.
Type refinement is part of a bigger topic in class-based languages: expressing and
preserving type recursion and (nominal/structural) subtyping at the same time.

One famous attempt in this direction is provided by
\emph{MyType}~\cite{bruce1994paradigmatic}, representing the type of
\textbf{this}, changing its meaning along with inheritance. However when
invoking a method with MyType in a parameter position, the exact type of the
receiver must be known.  This is a big limitation in class based object oriented
programming, and is exasperated by the interface-based programming we propose: no
type is ever going to be exact since classes are not explicitly used. A recent
article~\cite{Saito2013933} lights up this topic, proposing two
new features: exact statements and nonheritable methods. Both are
related to our work: 1) any method generated inside the \Q@of@ method is indeed
non-inheritable since there is no class name to extend from; 2) exact
statements (a form of wild-card capture on the exact run-time type) could
capture the ``exact type'' of an object even in a class-less
environment.
Admittedly, MyType greatly enhances the expressiveness and extensibility of
object-oriented programming languages. Object interfaces use covariant-return types to
simulate some uses of MyType. But this approach only works for refining
return types, whereas MyType is more general, as it also works for
parameter types. Our approach to covariantly refine state can recover
some of the additional expressiveness of MyType. As illustrated with our examples and
case studies, object interfaces are still very useful in many
practical applications, yet they do not require additional
complexity from the type system.

>>>>

\paragraph{Models of \objectoriented programming}
There are traditionally two big styles of OOP: \classbased and
\prototypebased. The \classbased tradition started with Simula~\cite{Dahl66simula},
and is the dominant style of OOP today. Major languages such
as Java, C\#, C++ or Scala favour a \classbased OOP style. Very often
\classbased languages are statically typed and they have interfaces or
some other similar mechanisms such as traits. However, in contrast to IB
programming, interfaces or traits in CB programming are not meant to
replace classes, but rather, they complement classes. Normally
interfaces or traits define the signatures of the methods and
\emph{may} have some behaviour (such as in Java 8 interfaces or Scala
traits). However classes are still responsible for object
instantiation and handling state. \Prototypebased languages started
with Self~\cite{Ungar87self}, and are still widely used today via languages like
JavaScript. Frequently \prototypebased languages are dynamically
typed. IB programming is clearly closer to CB programming. However the
key difference is that classes are not needed in IB
programming. Object interfaces recover the responsibility of object
instantiation, and abstract state operations provide a way to deal
with state. A primary motivation of IB programming is to improve
modularity and flexibility. Central to these goals are the
more expressive and automated forms of type-refinement, and support
for multiple interface inheritance.

\paragraph{Multiple inheritance in \objectoriented languages}
Many authors have argued in favor or against multiple inheritance.
Multiple \emph{implementation}
inheritance is very expressive, but difficult to model and implement, and can
cause difficulties (including the famous diamond (fork-join)
problem~\cite{bracha90mixin,Sak89dis}, conflicting methods, etc.) in reasoning about
programs. To conciliate the need for expressive power and simplicity, many
models have been proposed, including C++ virtual inheritance,
mixins~\cite{bracha90mixin}, traits~\cite{scharli03traits}, and hybrid models
such as CZ~\cite{malayeri2009cz}.  They provide novel programming architecture
models in the OO paradigm. In terms of restrictions set on these models, C++
virtual inheritance aims at a relative general model; the mixin model adds some
restrictions; and the trait model is the most restricted one (excluding
state, instantiation, etc).

@@

\paragraph{ThisType and MyType}
%\marco{ this section is unfair. There are unsolved problems with ThisType in negative positions, namely it breaks subtyping. The most common solution is to allow calling some methods only when the exact type is known. This demolishes most advantages of interfaces.}
Object interfaces support automatic type-refinement.
Type refinement is part of a bigger topic in \classbased languages: expressing and
preserving type recursion and (nominal/structural) subtyping at the same time.

One famous attempt in this direction is provided by
\emph{MyType}~\cite{bruce1994paradigmatic}, representing the type of
\textbf{this}, changing its meaning along with inheritance. However when
invoking a method with MyType in a parameter position, the exact type of the
receiver must be known.  This is a big limitation in \classbased \objectoriented
programming, and is exasperated by the \interfacebased programming we propose: no
type is ever going to be exact since classes are not explicitly used. A recent
article~\cite{Saito2013933} lights up this topic, proposing two
new features: exact statements and nonheritable methods. Both are
related to our work: 1) any method generated inside the \Q@of@ method is indeed
non-inheritable since there is no class name to extend from; 2) exact
statements (a form of wild-card capture on the exact run-time type) could
capture the ``exact type'' of an object even in a class-less
environment.
Admittedly, MyType greatly enhances the expressiveness and extensibility of
\objectoriented programming languages. Object interfaces use covariant-return types to
simulate some uses of MyType. But this approach only works for refining
return types, whereas MyType is more general, as it also works for
parameter types. Our approach to covariantly refine state can recover
some of the additional expressiveness of MyType. As illustrated with our examples and
case studies, object interfaces are still very useful in many
practical applications, yet they do not require additional
complexity from the type system.

### 2016-03-24 OOPSLA16/sections/Translation.tex
Spellcheck.
please use these commands in further editions to unify hyphen usage: \objectoriented \interfacebased \classbased \prototypebased ... you may refer to file predef/def.tex

Figure~\ref{figure:ofmethod} 
defines the $\ofMethod$  function, that generates the static method $\QM{of}$ as an
object factory.
It detects all the field methods of $\C_0$ and use them to synthesize
its arguments.
The return statement instantiates an anonymous class which
generates the needed getters, fluent setters and with-methods.
The right column first point collects the getter methods,
the second and third point generate implementations for \texttt{with-} methods
if needed; similarly, the fourth and fifth point generate fluent setters if needed.

@@

\paragraph{Type preservation}
Note that we preferred to introduce self, client and heir coherence instead of referring to conventional
type preservation theorems. The reason is
to better model how our approach behaves in a object oriented software ecosystem with inheritance,
where only some units may be translated/expanded.
Note inheritance's crucial influence in heir coherence.
Our formulation of client coherence allows us to discuss about intermediate
stages where only some code units are translated/expanded. Conventional 
type preservation refers only to completely translated program.
Our coherence guarantees mean that developers and designers of Java libraries and frameworks
can start using IB (and our \mixin annotation) in the evolution of their products
and still retain backward compatibility with their clients.

>>>>

Figure~\ref{figure:ofmethod}
defines the $\ofMethod$  function, which generates the static method $\QM{of}$ as an
object factory.
It detects all the field methods of $\C_0$ and use them to synthesize
its arguments.
The return statement instantiates an anonymous class which
generates the needed getters, fluent setters and with-methods.
The right column first point collects the getter methods,
the second and third point generate implementations for \texttt{with-} methods
if needed; similarly, the fourth and fifth point generate fluent setters if needed.

@@

\paragraph{Type preservation}
Note that we preferred to introduce self, client and heir coherence instead of referring to conventional
type preservation theorems. The reason is
to better model how our approach behaves in a \objectoriented software ecosystem with inheritance,
where only some units may be translated/expanded.
Note inheritance's crucial influence in heir coherence.
Our formulation of client coherence allows us to discuss about intermediate
stages where only some code units are translated/expanded. Conventional
type preservation refers only to completely translated program.
Our coherence guarantees mean that developers and designers of Java libraries and frameworks
can start using IB (and our \mixin annotation) in the evolution of their products
and still retain backward compatibility with their clients.
